---
title: "Getting started with Cacoa"
author: "Cacoa developers"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_depth: '3'
params:
  sim.file: ""      # path to the simulated dataset (test/sim_treatment.rds); empty = do not evaluate
  n.cores: 16
vignette: >
  %\VignetteIndexEntry{Getting started with Cacoa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# The walkthrough runs on a simulated dataset that is not shipped with the package (misc/notebook/prepare_sim_dataset.R
# builds it). The chunks are evaluated only when that file is given (misc/notebook/render_walkthrough.sh renders the
# executed Markdown copy in vignettes/); the vignette build evaluates nothing.
knitr::opts_chunk$set(eval = file.exists(params$sim.file), purl = FALSE, fig.width = 8, fig.height = 4.5, dpi = 110,
                      message = FALSE, warning = TRUE, comment = "")
```

Cacoa compares samples (patients, donors, replicates) across conditions. The analysis is organized around one
**model**: a formula with the condition of interest and the covariates to adjust for, and one **contrast**, the
comparison being made. The model is set once and every analysis uses it: expression shifts per cell type,
composition, differential expression, cell density and the cluster-free analyses.

The recommended order is

1. `Cacoa$new()` with the sample metadata.
2. `cao$checkDesign()`: is the planned comparison identifiable from the metadata alone?
3. `cao$screenCovariates()`: which covariates are associated with sample-level variation, in which cell types?
4. `cao$setModel()`: the model and the contrast used from here on.
5. `cao$estimateExpressionShiftMagnitudes()` and the other analyses.
6. `cao$checkSensitivity()`: does the result depend on the covariate set or on single samples?

This walkthrough uses a **simulated** dataset: 20 samples, 10 treated and 10 control, and 9 cell types. Expression
differences between the arms were planted with a known strength per cell type: strong in `IN-SST`, `IN-PV` and `L2/3`;
moderate in `OPC`, `L5/6-CC` and `AST-PP`; weak in `AST-FB`; none in `Microglia` and `Neu-NRGN`. The composition
is the same in both arms. A sample `age` was added as a covariate: the treated arm is older on average, but age has
no effect on expression in the simulation. The same code runs on any Conos or Seurat object with a sample metadata
table.

```{r libs}
library(cacoa)
library(ggplot2)
```

## Load data

The sample metadata is a data frame with one row per sample (row names are the sample names) and one column per
covariate.

```{r load}
sim <- readRDS(params$sim.file)
head(sim$sample.meta)
```

```{r construct, message=TRUE}
cao <- Cacoa$new(sim$con, sample.metadata = sim$sample.meta, cell.groups = sim$cell.groups, sample.per.cell = sim$sample.per.cell,
                 n.cores = params$n.cores)
```

The constructor reports which metadata columns can serve as covariates and which cannot (sample identifiers,
constant or mostly missing columns, too many levels). Options shared by all methods are set once; explicit arguments
of a method call still win. `plot.params` holds defaults for the embedding plots.

```{r options}
cao$setOptions(n.permutations = 499, seed = 1)
cao$plot.params <- list(size = 0.3, alpha = 0.3, font.size = 3)
```

## Check the design

Before fitting anything, the metadata alone tells whether the comparison can be adjusted for the other covariates:
associations between covariates, balance of the arms over them, aliasing, degrees of freedom and the number of
distinct permutations the design allows.

```{r check-design}
cao$checkDesign(test = "treatment")
```

```{r plot-design-balance, fig.width=4, fig.height=3.5}
cao$plotDesign("balance")
```

The treated arm is older on average, so age should be adjusted for; the check suggests comparing the adjusted and
the unadjusted result, which `checkSensitivity()` does below.

## Screen covariates

The screen tests every usable covariate against the sample-to-sample expression distances of each cell type, alone
(marginal) and adjusted for the other covariates (partial). The last column is a global test over all cell types.
The screen ends with a suggested model; it is not applied automatically.

```{r screen}
cao$screenCovariates(n.permutations = 199)
```

```{r plot-screen, fig.width=8, fig.height=3}
cao$plotCovariateScreen()
```

```{r plot-varpart, fig.width=8, fig.height=3.5}
cao$plotVariancePartition(c("treatment", "age"))
```

Treatment explains a large part of the between-sample variation in the planted cell types and none in `Microglia`
and `Neu-NRGN`; age explains nothing, as simulated.

## Set the model

`test` names the variable to test. For a two-level factor the contrast compares its two levels; the reference level
is chosen automatically (here `control`, matched by name) and `test = "treatment: control vs treated"` would reverse
it. The printout states the contrast, the adjustment set and the permutation scheme.

```{r set-model}
cao$setModel(~ treatment + age, test = "treatment")
```

## Expression shifts per cell type

For every cell type, each sample's cells are pooled into a pseudobulk profile and the sample-to-sample distances
are modelled with the formula above. The contrast gives three effects:

| effect | meaning |
|---|---|
| `shift` | the arms differ in a common direction, beyond the within-arm variability |
| `var`   | the treated arm is more (or less) heterogeneous than the control arm |
| `total` | treated samples are farther from control samples than control samples are from each other |

The effects are reported on a normalized scale: `shift` relative to the within-arm distances, `var` as the log2
ratio of the arm dispersions. P-values come from permutations and are adjusted across cell types
(Benjamini-Hochberg `padj`, and the familywise `p.fwer`); a global p-value per effect summarizes all cell types.

```{r shifts}
res <- cao$estimateExpressionShiftMagnitudes()
res
```

```{r plot-shifts, fig.width=9, fig.height=4}
cao$plotExpressionShiftMagnitudes()
```

Points are normalized effects with jackknife intervals; filled points are significant after adjustment across cell
types. The planted ordering is recovered: the three strong cell types on top, the two unchanged ones at zero.

### Behind one cell type

`plotShiftDetail()` shows the sample-level picture for one cell type: the samples in the adjusted distance space,
the distribution of pair distances within and between the arms, and the per-sample dispersion.

```{r plot-shift-detail, fig.width=12, fig.height=3.8}
cao$plotShiftDetail("IN-SST")
```

`plotSampleInfluence()` shows how much each sample changes the shift estimate of each cell type when it is left out,
in units of the jackknife standard error.

```{r plot-influence, fig.width=8, fig.height=4.5}
cao$plotSampleInfluence()
```

## Robustness

The shifts are re-estimated without adjustment, with the current model, and with the covariates proposed by the
screen. Per cell type, one verdict summarizes the comparison: robust, sign or significance depends on the model, the
estimate changes by more than half, or the result is driven by one sample.

```{r sensitivity}
cao$checkSensitivity(n.permutations = 199)
```

```{r plot-sensitivity, fig.width=9, fig.height=5}
cao$plotSensitivity()
```

Adjusting for age changes nothing here. The two cell types without a planted effect are flagged as driven by one
sample: their estimates are at zero and a single sample decides the sign.

## Sample structure

The sample distances of one cell type can be shown before and after the model adjustment, coloured by the tested
variable or by any covariate.

```{r plot-sample-distances, fig.width=9, fig.height=3.8}
cao$plotSampleDistances(cell.type = "IN-SST", color.by = "test")
```

## Differential expression and composition on the same model

The stored model is used by the per-cell-type differential expression, by the compositional analysis and by the
cell density analysis, and is recorded with each result.

```{r de}
de <- cao$estimateDEPerCellType(test = "limma-voom", min.cell.count = 5)
head(de[["IN-SST"]]$res[, c("Gene", "log2FoldChange", "pvalue", "padj", "CellFrac")])
```

```{r plot-n-de, fig.width=7, fig.height=3.5}
cao$plotNumberOfDEGenes()
```

Compositional changes are tested on the same contrast (cell-type loadings of the treated versus the control arm;
here none were planted):

```{r loadings, fig.width=5, fig.height=5}
cao$estimateCellLoadings()
cao$plotCellLoadings(show.pvals = FALSE)
```

## Cluster-free analyses

Cell density is compared between the arms in every neighbourhood of the cell graph, with the same adjustment, and
the cluster-free expression shifts test the contrast in every cell's neighbourhood.

```{r density, fig.width=10, fig.height=4.5}
cao$estimateCellDensity(method = "graph")
cao$estimateDiffCellDensity(type = "permutation")
cowplot::plot_grid(cao$plotEmbedding(color.by = "cell.groups"), cao$plotDiffCellDensity(), ncol = 2)
```

No density differences were simulated and none are found. The expression shifts, on the other hand, show the planted
cell types:

```{r cluster-free-shifts, fig.width=9, fig.height=4}
cao$estimateClusterFreeExpressionShifts(n.top.genes = 500, n.permutations = 99)
cao$plotClusterFreeExpressionShifts()
```

## Session info

```{r session-info}
sessionInfo()
```
