---
title: "Getting started with Cacoa"
author: '[Viktor Petukhov](https://github.com/VPetukhov/)'
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
  html_notebook:
    toc: yes
    toc_depth: '3'
    toc_float: yes
    df_print: paged
---

```{r libs, include=FALSE, message=FALSE}
library(sccore)
library(conos)
library(cacoa)
library(cowplot)
library(ggplot2)
```

## Load data

First, let's load a Conos object and create Cacoa from it:

```{r, message=FALSE, warning=FALSE}
con <- Conos$new(readRDS("../../cacoaAnalysis/data/PF/con.rds"))

cao <- cacoa::Cacoa$new(
  con, sample.groups=con$misc$sample_metadata$Diagnosis, 
  cell.groups=con$misc$cell_metadata$cellType,
  target.level='IPF', ref.level='Control', n.cores=45, verbose=FALSE
)

# set default plot parameters
cao$plot.params <- list(size=0.1, alpha=0.1, font.size=c(2, 3))
cao$plot.theme <- cao$plot.theme + theme(legend.background=element_blank())
```

## Cluster-based changes

The fastest way to visualize changes in the dataset is to show, what cell types 
changed their abundance and expression patterns.

### Estimate cluster-based changes

```{r, message=FALSE, warning=FALSE}
cao$estimateCellLoadings()
cao$estimateExpressionShiftMagnitudes()
```

### Plot compositional changes

```{r, fig.width=5, fig.height=6}
cao$plotCellLoadings(show.pvals=FALSE)
```

The red line here shows statistical significance. So we can see that Basal, KRT and 
SCGB3A2+ cell types are over-represented in IPF, while cell types like AT*, Monocytes
and other have higher abundance in Control.

### Plot expression changes

```{r, fig.width=6, fig.height=5, warning=FALSE, message=FALSE}
cao$plotExpressionShiftMagnitudes()
```

Here, y-axis shows magnitude of changes, while asterisks on top of bars show their significance.
We can see that AT2 cells show both expression and composition changes.

## Sample structure

Knowing sample metadata we can see, what factors affect sample differences:

```{r}
# sample_meta is a list or data.frame with metadata per sample
sample_meta <- con$misc$sample_metadata
head(as.data.frame(sample_meta))
```

This function shows warning if metadata significantly affects results:

```{r}
pvals <- cao$estimateMetadataSeparation(sample.meta=sample_meta)$padjust
sort(pvals)
```

Now we can show sample expression structure colored by one of the significant factors:

```{r, fig.width=4, fig.height=4}
cao$plotSampleDistances(
  space="expression.shifts",
  sample.colors=sample_meta$Sample_Source, legend.position=c(0, 1), font.size=2
)
```

The same can be done for compostional structure of samples: 

```{r, fig.width=4, fig.height=4}
cao$plotSampleDistances(
  space="coda",
  sample.colors=sample_meta$Sample_Source, legend.position=c(0, 1), font.size=2
)
```

## Functional interpretation

Estimate DE and GSEA:

```{r}
cao$estimateDEPerCellType(independent.filtering=TRUE, test='DESeq2.Wald', verbose=FALSE)
cao$estimateOntology(type="GSEA", org.db=org.Hs.eg.db::org.Hs.eg.db, verbose=FALSE, n.cores=1)
```

A quick way to look how many significant genes there are per cell type is to show a panel of volcano plots:

```{r, fig.width=10, fig.height=12, warning=FALSE, message=FALSE}
cao$plotVolcano(xlim=c(-3, 3), ylim=c(0, 3.5), lf.cutoff=1)
```

To better visualize GOs, we have a function that collapses ontologies with similar
enriched genes or similar enrichment pattern:

```{r, fig.width=9, fig.height=12}
cao$plotOntologyHeatmapCollapsed(
  name="GSEA", genes="up", n=50, clust.method="ward.D", size.range=c(1, 4)
)
```

You can also plot GO terms without collapsing them by enrichment patterns.
Usually, it takes hundreds of rows, so to make it readable you can focus on the
process of interest by filtering them by description:

```{r, fig.width=9, fig.height=4}
cao$plotOntologyHeatmap(
  name="GSEA", genes="up", description.regex="extracellular|matrix"
)
```

## Cluster-free

### Compositional changes

Estimate:

```{r, message=FALSE, warning=FALSE}
cao$estimateCellDensity(method='graph')
cao$estimateDiffCellDensity(type='wilcox')
```

Plot:

```{r, fig.width=8, fig.height=4, message=FALSE, warning=FALSE}
plot_grid(
  cao$plotEmbedding(color.by='cell.groups'), 
  cao$plotDiffCellDensity(legend.position=c(0, 1)), 
  ncol=2
)
```

### Expression changes

Estimate:

```{r}
cao$estimateClusterFreeExpressionShifts(
  gene.selection="expression", min.n.between=3, min.n.within=3
)
```

Plot:

```{r, fig.width=8, fig.height=4}
cao$plotClusterFreeExpressionShifts(legend.position=c(0, 1), font.size=2)
```

## Cluster-free DE

Estimate DE:

```{r}
cao$estimateClusterFreeDE(
  n.top.genes=1000, min.expr.frac=0.01, adjust.pvalues=TRUE, smooth=TRUE, 
  verbose=TRUE
)
```

Estimate gene programs:

```{r, warning=FALSE, message=FALSE}
cao$estimateGenePrograms(method="leiden", z.adj=TRUE, smooth=FALSE)
```

Plot gene programs:

```{r, fig.width=12, fig.height=12}
cao$plotGeneProgramScores(
  legend.position=c(0, 1), plot.na=FALSE, 
  adj.list=theme(legend.key.width=unit(8, "pt"), legend.key.height=unit(12, "pt"))
)
```

Plot genes from one program:

```{r, fig.width=9, fig.height=9}
plot_grid(plotlist=cao$plotGeneProgramGenes(
  program.id=4, max.genes=9, plot.na=FALSE, legend.position=c(0, 1)
), ncol=3)
```

## Session Info

```{r, echo=FALSE}
data.frame(value=unlist(sessioninfo::platform_info()))
tibble::as_tibble(
  sessioninfo::package_info()
)[c('package', 'loadedversion', 'date', 'source')]
```

