---
title: "Getting started with Cacoa"
author: "Cacoa developers"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_depth: '3'
params:
  sim.file: ""      # path to the simulated dataset; empty = do not evaluate
  n.cores: 16
vignette: >
  %\VignetteIndexEntry{Getting started with Cacoa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# The walkthrough runs on a simulated dataset (720 MB) that is not shipped with the package. The chunks are
# evaluated only when that file is present (misc/notebook/render_walkthrough.sh renders the executed Markdown
# copy in vignettes/); the vignette build evaluates nothing.
knitr::opts_chunk$set(eval = file.exists(params$sim.file), purl = FALSE, fig.width = 8, fig.height = 4.5, dpi = 110,
                      message = FALSE, warning = TRUE, comment = "")
```

Cacoa compares samples (patients, donors, replicates) across conditions. One **model** is set once and used by every
analysis: expression shifts per cell type, composition, differential expression, cell density and the cluster-free
analyses.

The recommended order is

1. `Cacoa$new()` with the sample metadata. No model is needed yet; a summary of the metadata is printed.
2. `cao$checkDesign()`, `cao$plotDesign()`: confounding, balance, degrees of freedom and permutation feasibility, from the metadata alone.
3. `cao$screenCovariates()`, `cao$plotCovariateScreen()`: which covariates are associated with sample-level variation, and in which cell types (exploratory).
4. `cao$setModel(~ condition + batch, test = "condition")`: the model used from here on.
5. `cao$estimateExpressionShiftMagnitudes()`, `cao$plotExpressionShiftMagnitudes()` and the other analyses.
6. `cao$checkSensitivity()`, `cao$plotSensitivity()`: is the result robust to the covariate set and to single samples?

This walkthrough runs the workflow on a **simulated** dataset: a Conos object with 40 samples (4 groups x 2 batches,
5 samples each) and 9 cell types of 500 cells. Expression differences between `Group2` and `Group1` were planted with
a known strength per cell type: strong in `IN-SST`, `IN-PV` and `L2/3`; moderate in `OPC`, `L5/6-CC` and `AST-PP`;
weak in `AST-FB`; none in `Microglia` and `Neu-NRGN`. The object is not shipped with the package (720 MB); the
executed copy of this document is provided for reading, with its outputs. The same code runs on any Conos or Seurat
object with a sample metadata table.

```{r libs}
library(conos)
library(cacoa)
library(ggplot2)
library(cowplot)
```

## Load data

`sample.metadata` is a data frame with one row per sample (row names are the sample names) and one column per
covariate. Here it has the group, the batch and two technical columns from the simulation. A random `age` column is
added to show how an unrelated covariate is treated.

```{r load}
sim <- readRDS(params$sim.file)$with_batch              # a Cacoa object from an earlier version
con <- sim$data.object                                 # the Conos object inside it
sample.meta <- con$misc$sample_meta
set.seed(1); sample.meta$age <- round(rnorm(nrow(sample.meta), 55, 8))   # unrelated to the groups, for illustration
head(sample.meta)
```

```{r construct, message=TRUE}
cao <- Cacoa$new(con, sample.metadata = sample.meta, cell.groups = sim$cell.groups, sample.per.cell = sim$sample.per.cell,
                 n.cores = params$n.cores)
cao$plot.params <- list(size = 0.3, alpha = 0.3, font.size = c(2, 3))
```

The constructor prints which metadata columns can be used as covariates and which cannot (ID-like, constant, mostly
missing, too many levels). The previous two-group constructor (`sample.groups`, `ref.level`, `target.level`) still
works and is translated into a `condition` column and `test = "condition"`.

Options shared by all methods are set once; explicit arguments of a method call still win.

```{r options}
cao$setOptions(n.permutations = 499, seed = 1)
```

## Check the design

Before fitting anything, the metadata alone tells whether the planned comparison can be adjusted for the other
covariates: associations between covariates (bias-corrected Cramer's V, correlation ratio, Spearman correlation),
balance of the groups over the covariates, aliasing, degrees of freedom and how many distinct permutations the design
allows.

```{r check-design}
chk <- cao$checkDesign(test = "Group")       # the report is printed; `chk` keeps the tables
```

```{r plot-design-assoc, fig.width=5, fig.height=4}
cao$plotDesign("associations")
```

```{r plot-design-balance, fig.width=8, fig.height=3.5}
cao$plotDesign("balance")
```

The report lists errors (a covariate that fully determines the condition), warnings (strongly associated covariates,
small groups, few distinct permutations) and notes, each with a suggestion. Here the groups are balanced over the
batches; the simulation's library-size column is strongly associated with the group, so the report suggests comparing
models with and without it (`cao$checkSensitivity()`), which is done below.

## Screen covariates

The screen tests every usable covariate against the sample-to-sample expression distances of each cell type,
**marginally** (alone) and **partially** (adjusted for all other covariates). A covariate that is significant only
marginally is explained by the others (shown as a ring in the plot). The last column is a global test over all cell
types. The screen ends with a suggested model; it is not applied automatically.

```{r screen}
scr <- cao$screenCovariates(n.permutations = 199)   # `scr$table` holds the per-covariate results
```

```{r plot-screen, fig.width=8, fig.height=4}
cao$plotCovariateScreen()
```

```{r plot-screen-summary, fig.width=7, fig.height=3.5}
cao$plotCovariateSummary()
```

Variance partition: how much of the sample-level variation is explained uniquely by each covariate, shared between
them, or left unexplained.

```{r plot-varpart, fig.width=8, fig.height=3.5}
cao$plotVariancePartition(c("Group", "Batch"))
```

## Set the model

`test` accepts a variable name (two levels: a contrast with an automatically chosen reference; numeric: a slope per
unit; a factor with more levels needs the comparison, e.g. `"Group: Group2 vs Group1"`, or `"Group: all"` for a
whole-factor test), an explicit comparison such as `"Group: Group2 vs Group1"`, several variables, `"all"`, or a
structured contrast. The printout states the reference level and how it was chosen, the adjustment set, the
permutation scheme and the number of distinct permutations.

Here two tests are set: the planted contrast `Group2 vs Group1`, and the whole-factor test of `Group` over all four
groups. Both are adjusted for `Batch`.

```{r set-model}
cao$setModel(~ Group + Batch, test = c("Group: Group2 vs Group1", "Group: all"))
```

## Expression shifts per cell type

The expression shift is estimated with an individual-level model fitted to the sample-sample distances of each cell
type. Three effects are reported for a contrast:

| effect | meaning |
|---|---|
| `shift` | the groups differ in a common direction, beyond within-group variability |
| `var`   | the target group is more (or less) heterogeneous than the reference group |
| `total` | target samples are farther from reference samples than reference samples are from each other (`shift + var/2`) |

For a whole-factor test the effects are `location` (any group differs) and `dispersion` (the groups differ in
heterogeneity). P-values come from permutations (block randomization within batches here); they are adjusted across
cell types by Benjamini-Hochberg (`padj`) and by the max-statistic (`p.fwer`), and a global p-value per test is
reported.

```{r shifts}
res <- cao$estimateExpressionShiftMagnitudes()
subset(res$results, effect == "shift")[, c("celltype", "estimate", "estimate.norm", "se.jk", "p", "padj", "p.fwer", "n")]
```

```{r shifts-global}
res$global
```

```{r plot-shifts, fig.width=9, fig.height=4.5}
cao$plotExpressionShiftMagnitudes()
```

Points are normalized effects with jackknife intervals; filled points are significant after Benjamini-Hochberg
adjustment across cell types. The subtitle records the test, the adjustment set and the permutation scheme. The
planted strong cell types come out on top and the two unchanged ones at the bottom.

The second test (all four groups) is plotted by name.

```{r plot-shifts-term, fig.width=9, fig.height=4.5}
cao$plotExpressionShiftMagnitudes(test = "Group")
```

### Behind one cell type

`plotShiftDetail()` shows the sample-level picture: the distance distribution within and between groups, and the
samples in the adjusted distance space.

```{r plot-shift-detail, fig.width=10, fig.height=4}
cao$plotShiftDetail("IN-SST")
```

`plotSampleInfluence()` shows how much each sample changes the shift estimate when it is left out.

```{r plot-influence, fig.width=9, fig.height=4.5}
cao$plotSampleInfluence()
```

## Robustness

The expression shifts are re-estimated without adjustment, with the current model, without each covariate, and with
the top covariates proposed by the screen. Per cell type, one verdict summarizes the result: robust, sign or
significance sensitive to the adjustment set, the estimate changes by more than half, or the result is driven by one
sample.

```{r sensitivity}
sens <- cao$checkSensitivity(n.permutations = 199)
```

```{r plot-sensitivity, fig.width=9, fig.height=5}
cao$plotSensitivity()
```

## Sample structure

The sample distances of one cell type (or of all cell types jointly) can be shown as is, or after removing the part
explained by a covariate (`adjust.for`).

```{r plot-sample-distances, fig.width=10, fig.height=4}
plot_grid(
  cao$plotSampleDistances(space = "expression.shifts", cell.type = "IN-SST", color.by = "test", values = "unadjusted"),
  cao$plotSampleDistances(space = "expression.shifts", cell.type = "IN-SST", color.by = "Batch", adjust.for = ~ Group),
  ncol = 2)
```

## Differential expression and composition on the same model

The stored model (formula, test, reference level) is used by the per-cell-type differential expression, by the
compositional analysis and by the cell density analysis, and is recorded with each result. Whole-factor tests are
routed to the corresponding multi-group tests (DESeq2 LRT, edgeR/limma F).

```{r de}
de <- cao$estimateDEPerCellType(test = "limma-voom", min.cell.count = 5)
sapply(de, function(d) sum(d$res$padj < 0.05, na.rm = TRUE))             # DE genes per cell type
```

The top genes of one cell type (`cao$plotVolcano()` draws them when the EnhancedVolcano package is installed):

```{r de-top}
r <- de[["IN-SST"]]$res; r <- r[order(r$padj), intersect(c("Gene", "log2FoldChange", "stat", "pvalue", "padj", "CellFrac"), names(r))]
head(r, 8)
```

```{r loadings, fig.width=5, fig.height=5}
cao$estimateCellLoadings()
cao$plotCellLoadings(show.pvals = FALSE)
```

## Cluster-free analyses

Cell density is compared between the two groups using the regression weights of the contrast, so adjusted covariates
are accounted for. Cluster-free expression shifts test the same contrast in every cell's neighbourhood, sharing the
permutations across cells and adjusting the z-scores by the maximum statistic.

```{r density, fig.width=10, fig.height=4.5}
cao$estimateCellDensity(method = "graph")
cao$estimateDiffCellDensity(type = "permutation")
plot_grid(cao$plotEmbedding(color.by = "cell.groups"), cao$plotDiffCellDensity(), ncol = 2)
```

```{r cluster-free-shifts, fig.width=5.5, fig.height=4.5}
cao$estimateClusterFreeExpressionShifts(n.top.genes = 500, n.permutations = 99)
cao$plotClusterFreeExpressionShifts(font.size = 2)
```

## Session info

```{r session-info}
sessionInfo()
```
