---
title: "Getting started with Cacoa"
author: "Cacoa developers"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Getting started with Cacoa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, eval=FALSE, purl=FALSE}
# The walkthrough uses a Conos object of a pulmonary fibrosis cohort that is not shipped with the package,
# so the chunks are not evaluated when the vignette is built.
knitr::opts_chunk$set(eval = FALSE, purl = FALSE)
```

```{r libs, message=FALSE, eval=FALSE, purl=FALSE}
library(conos)
library(cacoa)
library(cowplot)
library(ggplot2)
```

## The workflow in one glance

Cacoa compares samples (patients, donors, replicates) across conditions. One **model** is set once and used by
every analysis: expression shifts per cell type, composition, differential expression, cell density and the
cluster-free analyses. The recommended order is

1. `Cacoa$new()` with the sample metadata: no model is needed yet; a summary of the metadata is printed.
2. `cao$checkDesign()`, `cao$plotDesign()`: confounding, balance, degrees of freedom, permutation feasibility,
   from the metadata alone.
3. `cao$screenCovariates()`, `cao$plotCovariateScreen()`: which covariates are associated with sample-level
   variation, in which cell types (exploratory).
4. `cao$setModel(~ condition + batch, test = "condition")`: the model used from here on.
5. `cao$estimateExpressionShiftMagnitudes()`, `cao$plotExpressionShiftMagnitudes()` and the other analyses.
6. `cao$checkSensitivity()`, `cao$plotSensitivity()`: is the result robust to the covariate set and to single samples?

## Load data

```{r, eval=FALSE, purl=FALSE}
con <- Conos$new(readRDS("../../cacoaAnalysis/data/PF/con.rds"))
sample.meta <- as.data.frame(con$misc$sample_metadata)   # rows = samples, columns = covariates
rownames(sample.meta) <- sample.meta$Sample_Name

cao <- Cacoa$new(con, sample.metadata = sample.meta, cell.groups = con$misc$cell_metadata$cellType, n.cores = 45)

cao$plot.params <- list(size = 0.1, alpha = 0.1, font.size = c(2, 3))
cao$plot.theme <- cao$plot.theme + theme(legend.background = element_blank())
```

The constructor prints which metadata columns can be used as covariates and which cannot (ID-like, constant,
mostly missing, too many levels). The previous two-group constructor (`sample.groups`, `ref.level`,
`target.level`) still works and is translated into a `condition` column and `test = "condition"`.

## Check the design

```{r, eval=FALSE, purl=FALSE}
cao$checkDesign(test = "Diagnosis")
cao$plotDesign("associations")
cao$plotDesign("balance")
```

The report lists errors (a covariate that fully determines the condition), warnings (strongly associated
covariates, small groups, few distinct permutations) and notes, each with a suggestion.

## Screen covariates

```{r, eval=FALSE, purl=FALSE}
cao$screenCovariates()            # all usable covariates, marginal and partial, per cell type
cao$plotCovariateScreen()         # dot: FDR < 0.05; ring: significant only marginally (explained by other covariates)
cao$plotCovariateSummary()
cao$plotVariancePartition(c("Diagnosis", "Sample_Source"))
```

The screen ends with a suggested model. It is not applied automatically.

## Set the model

```{r, eval=FALSE, purl=FALSE}
cao$setModel(~ Diagnosis + Sample_Source, test = "Diagnosis")
```

`test` accepts a variable name (two levels: a contrast with an automatically chosen reference; numeric: a slope per
unit; a factor with more levels needs the comparison, e.g. `"Group: G2 vs G1"`, or `"Group: all"` for a whole-factor
test), an explicit comparison such as `"Diagnosis: IPF vs Control"`, several variables, `"all"`, or a structured
contrast. The printout states the reference level and how it was chosen, the
adjustment set, the permutation scheme and the number of distinct permutations.

Options shared by all methods are set once:

```{r, eval=FALSE, purl=FALSE}
cao$setOptions(n.permutations = 1999, seed = 1, n.cores = 45)
```

## Cluster-based changes

```{r, eval=FALSE, purl=FALSE}
cao$estimateCellLoadings()
cao$estimateExpressionShiftMagnitudes()
```

```{r, fig.width=5, fig.height=6, eval=FALSE, purl=FALSE}
cao$plotCellLoadings(show.pvals = FALSE)
```

```{r, fig.width=9, fig.height=5, eval=FALSE, purl=FALSE}
cao$plotExpressionShiftMagnitudes()
```

Three effects are reported per cell type:

| effect | meaning |
|---|---|
| `shift` | the groups differ in a common direction, beyond within-group variability |
| `var`   | the target group is more (or less) heterogeneous than the reference group |
| `total` | target samples are farther from reference samples than reference samples are from each other (`shift + var/2`) |

Points are normalized effects with jackknife intervals; filled points are significant after Benjamini-Hochberg
adjustment across cell types. The subtitle records the test, the adjustment set and the permutation scheme.
`cao$plotShiftDetail("AT2")` shows the sample-level picture behind one cell type, `cao$plotSampleInfluence()` the
effect of leaving each sample out.

## Robustness

```{r, eval=FALSE, purl=FALSE}
cao$checkSensitivity()
cao$plotSensitivity()
```

The expression shifts are re-estimated without adjustment, with the current model, without each covariate, and
with the top covariates proposed by the screen. One verdict per cell type summarizes the result.

## Sample structure

```{r, fig.width=8, fig.height=4, eval=FALSE, purl=FALSE}
cao$plotSampleDistances(space = "expression.shifts", color.by = "test", values = "both")
cao$plotSampleDistances(space = "expression.shifts", color.by = "Sample_Source", adjust.for = ~ Diagnosis)
```

## Functional interpretation

```{r, eval=FALSE, purl=FALSE}
cao$estimateDEPerCellType(independent.filtering = TRUE, test = "DESeq2.Wald")
cao$estimateOntology(type = "GSEA", org.db = org.Hs.eg.db::org.Hs.eg.db, n.cores = 1)
```

```{r, fig.width=10, fig.height=12, eval=FALSE, purl=FALSE}
cao$plotVolcano(xlim = c(-3, 3), ylim = c(0, 3.5), lf.cutoff = 1)
```

```{r, fig.width=9, fig.height=12, eval=FALSE, purl=FALSE}
cao$plotOntologyHeatmapCollapsed(name = "GSEA", genes = "up", n = 50, clust.method = "ward.D", size.range = c(1, 4))
```

## Cluster-free analyses

```{r, eval=FALSE, purl=FALSE}
cao$estimateCellDensity(method = "graph")
cao$estimateDiffCellDensity(type = "permutation")
cao$estimateClusterFreeExpressionShifts()
```

```{r, fig.width=8, fig.height=4, eval=FALSE, purl=FALSE}
plot_grid(cao$plotEmbedding(color.by = "cell.groups"), cao$plotDiffCellDensity(legend.position = c(0, 1)), ncol = 2)
cao$plotClusterFreeExpressionShifts(legend.position = c(0, 1), font.size = 2)
```

```{r, eval=FALSE, purl=FALSE}
cao$estimateClusterFreeDE(n.top.genes = 1000, min.expr.frac = 0.01, adjust.pvalues = TRUE, smooth = TRUE)
cao$estimateGenePrograms(method = "leiden", z.adj = TRUE, smooth = FALSE)
```

## Session Info

```{r, echo=FALSE, eval=FALSE, purl=FALSE}
sessionInfo()
```
